A Plot is Worth a Thousand Tests: Assessing Residual Diagnostics with the Lineup Protocol

### 2023-04-22 paper.Rmd
up to section 4

The power of a model misspecification test is the probability that $H_0$ is rejected given the regression model is misspecified in a specific way. It is an important indicator
when one is concerned about whether model assumptions have been violated. Although in practice, one might be more interested in knowing how much the residuals deviate from the model assumptions, and whether this deviation is of practical significance.

The power of a conventional hypothesis test is affected by both the true parameter $\boldsymbol{\theta}$ and the sample size $n$. These two can be quantified in terms of effect size $E$ to measure the strength of the residual departures from the model assumptions. Details about the effect size is provided in Section \ref{effect-size} after the introduction of the simulation model used in our human subject experiment. The theoretical power of a test is sometimes not a trival solution, but it can be estimated if the data generating process is known. We use a predefined model to generate a large set of simulated data under different effect sizes, and record if the conventional test rejects $H_0$. The probability of the conventional test rejects $H_0$ is then fitted by a logistic regression formulated as

@@

\noindent where $\Lambda(.)$ is the standard logistic function given as $\Lambda(z) = exp(z)/(1+exp(z))$. The effect size $E$ is the only predictor and the intercept is fixed to $log(0.05/0.95)$ so that $\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05$, which is the desired significance level.

The power of a visual test on the other hand, may additionally depend on the ability of the particular subject, as the skill of the individual may affect the number of observers who identify the data plot from the lineup [@majumder_validation_2013]. To address this issue, @majumder_validation_2013 models the probability of a subject $j$ correctly picking the data plot from a lineup $l$ using a mixed-effect logistic regression, with subjects treated as random effect. Then, the estimated power of a visual test evaluated by a single subject is the predicted value obtained from the mixed effects model. However, this mixed effects model does not work with scenario where subjects are asked to select one or more most different plots. In this scenario, having the probability of a subject $j$ correctly picking the data plot from a lineup $l$ is insufficient to determine the power of a visual test because it does not provide information about the number of selections made by the subject for the calculation of the $p$-value (See Equation \ref{eq:pvalue-beta-binomial-approx}). Therefore, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. This assumption is not necessary true, but it helps simplifying the model structure, thereby obviate a costly large-scale experiment to estimate complex covariance matrices. The same model given in Equation \ref{eq:logistic-regression-1-1} is applied to model the power of a visual test.

>>>>

The power of a model misspecification test is the probability that $H_0$ is rejected given the regression model is misspecified in a specific way. It is an important indicator
when one is concerned about whether model assumptions have been violated. In practice, one might be more interested in knowing how much the residuals deviate from the model assumptions, and whether this deviation is of practical significance.

The power of a conventional hypothesis test is affected by both the true parameter $\boldsymbol{\theta}$ and the sample size $n$. These two can be quantified in terms of effect size $E$ to measure the strength of the residual departures from the model assumptions. Details about the effect size are provided in Section \ref{effect-size} after the introduction of the simulation model used in our experiment. The theoretical power of a test is sometimes not a trival solution, but it can be estimated if the data generating process is known. We use a predefined model to generate a large set of simulated data under different effect sizes, and record if the conventional test rejects $H_0$. The probability of the conventional test rejects $H_0$ is then fitted by a logistic regression formulated as

@@

\noindent where $\Lambda(.)$ is the standard logistic function given as $\Lambda(z) = exp(z)/(1+exp(z))$. The effect size $E$ is the only predictor and the intercept is fixed to $log(0.05/0.95)$ so that $\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05$, the desired significance level.

The power of a visual test on the other hand, may additionally depend on the ability of the particular subject, as the skill of each individual may affect the number of observers who identify the data plot from the lineup [@majumder_validation_2013]. To address this issue, @majumder_validation_2013 models the probability of subject $j$ correctly picking the data plot from lineup $l$ using a mixed-effect logistic regression, with subjects treated as random effect. Then, the estimated power of a visual test evaluated by a single subject is the predicted value obtained from the mixed effects model. However, this mixed effects model does not work with scenario where subjects are asked to select one or more most different plots. In this scenario, having the probability of a subject $j$ correctly picking the data plot from a lineup $l$ is insufficient to determine the power of a visual test because it does not provide information about the number of selections made by the subject for the calculation of the $p$-value (See Equation \ref{eq:pvalue-beta-binomial-approx}). Therefore, we directly estimate the probability of a lineup being rejected by assuming that individual skill has negligible effect on the variation of the power. This assumption is not necessary true, but it helps to simplify the model structure, thereby obviating a costly large-scale experiment to estimate complex covariance matrices. The same model given in Equation \ref{eq:logistic-regression-1-1} is applied to model the power of a visual test.

### 2023-04-22 paper.tex
up to section 4

The power of a model misspecification test is the probability that
\(H_0\) is rejected given the regression model is misspecified in a
specific way. It is an important indicator when one is concerned about
whether model assumptions have been violated. Although in practice, one
might be more interested in knowing how much the residuals deviate from
the model assumptions, and whether this deviation is of practical
significance.

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
effect size is provided in Section \ref{effect-size} after the
introduction of the simulation model used in our human subject
experiment. The theoretical power of a test is sometimes not a trival
solution, but it can be estimated if the data generating process is
known. We use a predefined model to generate a large set of simulated
data under different effect sizes, and record if the conventional test
rejects \(H_0\). The probability of the conventional test rejects
\(H_0\) is then fitted by a logistic regression formulated as

@@

\noindent where \(\Lambda(.)\) is the standard logistic function given
as \(\Lambda(z) = exp(z)/(1+exp(z))\). The effect size \(E\) is the only
predictor and the intercept is fixed to \(log(0.05/0.95)\) so that
\(\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05\), which is the desired
significance level.

The power of a visual test on the other hand, may additionally depend on
the ability of the particular subject, as the skill of the individual
may affect the number of observers who identify the data plot from the
lineup \citep{majumder_validation_2013}. To address this issue,
\citet{majumder_validation_2013} models the probability of a subject
\(j\) correctly picking the data plot from a lineup \(l\) using a
mixed-effect logistic regression, with subjects treated as random
effect. Then, the estimated power of a visual test evaluated by a single
subject is the predicted value obtained from the mixed effects model.
However, this mixed effects model does not work with scenario where
subjects are asked to select one or more most different plots. In this
scenario, having the probability of a subject \(j\) correctly picking
the data plot from a lineup \(l\) is insufficient to determine the power
of a visual test because it does not provide information about the
number of selections made by the subject for the calculation of the
\(p\)-value (See Equation \ref{eq:pvalue-beta-binomial-approx}).
Therefore, we directly estimate the probability of a lineup being
rejected by assuming that individual skill has negligible effect on the
variation of the power. This assumption is not necessary true, but it
helps simplifying the model structure, thereby obviate a costly
large-scale experiment to estimate complex covariance matrices. The same
model given in Equation \ref{eq:logistic-regression-1-1} is applied to
model the power of a visual test.

>>>>

The power of a model misspecification test is the probability that
\(H_0\) is rejected given the regression model is misspecified in a
specific way. It is an important indicator when one is concerned about
whether model assumptions have been violated. In practice, one might be
more interested in knowing how much the residuals deviate from the model
assumptions, and whether this deviation is of practical significance.

The power of a conventional hypothesis test is affected by both the true
parameter \(\boldsymbol{\theta}\) and the sample size \(n\). These two
can be quantified in terms of effect size \(E\) to measure the strength
of the residual departures from the model assumptions. Details about the
effect size are provided in Section \ref{effect-size} after the
introduction of the simulation model used in our experiment. The
theoretical power of a test is sometimes not a trival solution, but it
can be estimated if the data generating process is known. We use a
predefined model to generate a large set of simulated data under
different effect sizes, and record if the conventional test rejects
\(H_0\). The probability of the conventional test rejects \(H_0\) is
then fitted by a logistic regression formulated as

@@

\noindent where \(\Lambda(.)\) is the standard logistic function given
as \(\Lambda(z) = exp(z)/(1+exp(z))\). The effect size \(E\) is the only
predictor and the intercept is fixed to \(log(0.05/0.95)\) so that
\(\hat{Pr}(\text{reject}~H_0|H_1,E = 0) = 0.05\), the desired
significance level.

The power of a visual test on the other hand, may additionally depend on
the ability of the particular subject, as the skill of each individual
may affect the number of observers who identify the data plot from the
lineup \citep{majumder_validation_2013}. To address this issue,
\citet{majumder_validation_2013} models the probability of subject \(j\)
correctly picking the data plot from lineup \(l\) using a mixed-effect
logistic regression, with subjects treated as random effect. Then, the
estimated power of a visual test evaluated by a single subject is the
predicted value obtained from the mixed effects model. However, this
mixed effects model does not work with scenario where subjects are asked
to select one or more most different plots. In this scenario, having the
probability of a subject \(j\) correctly picking the data plot from a
lineup \(l\) is insufficient to determine the power of a visual test
because it does not provide information about the number of selections
made by the subject for the calculation of the \(p\)-value (See Equation
\ref{eq:pvalue-beta-binomial-approx}). Therefore, we directly estimate
the probability of a lineup being rejected by assuming that individual
skill has negligible effect on the variation of the power. This
assumption is not necessary true, but it helps to simplify the model
structure, thereby obviating a costly large-scale experiment to estimate
complex covariance matrices. The same model given in Equation
\ref{eq:logistic-regression-1-1} is applied to model the power of a
visual test.
